The spread of COVID-19 in a landlocked country the case of Luxembourg

### 2021-03-05 paper/paper.tex
updates before blog post

---
title: The spread of COVID-19 in a landlocked country the case of Luxembourg
authors:
  - name: Bruno Rodrigues
    thanks: This preprint was written during my free time as a private citizen, and reflects in no manner the views of the Ministry of Higher Education and Research, nor the Government of the Grand-Duchy of Luxembourg.
    department: Statistics department
    affiliation: Ministry of Higher Education and Research, Grand-Duchy of Luxembourg
    location: Luxembourg, 18-20, Montée de la Pétrusse
    email: bruno.rodrigues@mesr.etat.lu
abstract: |
  Enter the text of your abstract here.
keywords:
  - blah
  - blee
  - bloo
  - these are optional and can be removed
bibliography: references.bib
biblio-style: unsrt
output: 
  rticles::arxiv_article:
    keep_tex: true
---

# Introduction

@@

In this article, I posit the following hypothesis: due to its quite unique characteristics, 
the spread of COVID-19 in a landlocked country like Luxembourg is the exact opposite of the 
spread of COVID-19 that can be observed on an island country such as New Zealand, or Madagascar.
A landlocked country like Luxembourg, which is furthermore highly dependent on foreign workers, 
has many more difficulties to control the spread of COVID-19 within its borders. Unlike an island
country, a landlocked country that is highly tied to its neighbours cannot simply close its borders
and put a very hard lockdown in place to control the pandemic. Or if the landlocked country
does that, as soon as it opens its borders, the disease will start spreading again.
To illustrate this idea, I will discuss how COVID-19 starting spreading, but not only within the 
borders of Luxembourg, but rather within the so-called Greater Region. 
The Greater Region *a space for cross-border cooperation in the heart of Europe* and 
is composed of the Grand-Duchy of Luxembourg, two Belgian Provinces, two French Départements and 
two German Bundesländer.

@@

The approach I will be using in this paper is thus as follows: I will train a machine learning model to predict the 
spread of COVID-19 in Luxembourg using openly available data on the weekly positive cases of 
COVID-19. However, because of the very tight economic and social integration of Luxembourg to its 
neighbours I will use as features weekly positive cases in the border regions as well as 
Google Mobility data\footnote{\url{https://www.google.com/covid19/mobility/}} for Luxembourg
to proxy for hard, and soft, lockdowns. I will show that lags of weekly cases in the neighbouring 
regions predict cases for Luxembourg. The end goal however, is *not* to build a model to predict
how many weekly positive cases will be detected in Luxembourg. This would be a fools errand; 
in order to predict the future, the future has to look like the past, but in the case of this pandemic
there is absolutely no guarantee that the future will look like the past, and there are many reasons
for this. First of all, people are constantly adapting their behaviour, and public health policies
are also constantly being tuned, and getting sometimes more restrictive, sometimes more relaxed.
Secondly, vaccines have started being administrated and it would be impossible to predict the effect
on weekly positive cases using the approach I'm using. Finally, there's also the threat of the 
variants. Here again, it is impossible to predict which new variants could arise and how much more
contagious -and deadly- these could be.
So then, why bother with this paper? The end goal is not prediction, but explainability. Once the 
model is trained, I will use explainability methods to show which variables, and their interaction
with each other, predict positive cases for Luxembourg. This will be a clear illustration of the
hypothesis that I posited at the beginning; that a landlocked country like Luxembourg which is very
tightly economically and socially integrated with its neighbours cannot fight a pandemic on its own,
but must cooperate with its neighbours. This argument can also be applied to any other country in
the same situation as Luxembourg or even to the constituent states of a federal nation. Unfortunately, 
the virus does not respect the borders of sovereign nations.

# The COVID-19 pandemic in the Greater Region

COVID-19 started to spread in the Greater Region around end of February. Two french *départements*,
the *Bas-Rhin* and the *Haut-Rhin*, which together form the historical region of Alsace, were
hit very hard. A religious gathering in the *Haut-Rhin*, with more than 2000 worshipers is the very
likely start of the spread in Alsace, and then to the neighbouring Lorraine administrative region. 
Alsace is part of the *Grand Est* region, as is Lorraine. Two *départements* 
of *Lorraine* are part of the Greater Region, and the disease spread there quickly as well.

@@

```{r epid_curves, fig.cap = "Test"}
tar_read(epid_curves)
```

In Figure \@ref(fig:epid_curves)

```{r, out.width = "100%"}
tar_read(epidem_map)$map_first_wave
```

```{r, out.width = "100%"}
tar_read(epidem_map)$map_second_wave
```

# The data used

Data on positive cases from the regions of the Greater Region was collected through each of the 
countries' open data portal. The levels of detail were heterogeneous, with Belgium and Germany
providing a high level of detail (daily cases by sex, age group, Province in the case of Belgium, and
Land- and Stadtkreise in the case of Germany), followed by France (daily cases by department and
age group), with Luxembourg providing the least amount of details; only daily cases at the national
level. 
In order to simplify the process of getting the data from all these sources, we wrote an R package
called `{covidGrandeRegion}` which can be found on the following [github repository](https://github.com/b-rodrigues/covidGrandeRegion).
This R package provides several functions to download daily or weekly data, either for one single country or for
the whole of the Greater Region as well as a function to call an interactive map of the region
with a timeline, making it easy to visualise the spread of the disease through the region. It is also
possible to normalize the data by dividing the daily or weekly cases by the size of the population
in each sub-region.
Another variable that was included comes from the [Google Mobility website](https://www.google.com/covid19/mobility/).
This data shows on a daily basis how communities move since the start of the pandemic. This data
is used here as proxy for lockdowns. The plot below shows the daily percentage change in time 
spent at home in Luxembourg:

```{r}
tar_read(plot_mobility)
```

@@

```{r}
print_data <- tar_read(data_for_model) %>%
  mutate(across(where(is.numeric), ~round(., 0))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var="Variable names") %>%
  select(seq(1, 6)) %>%
  unite("First 5 observations", 2:6, sep = ", ")

kbl(print_data, booktabs = TRUE, position = "center")
```

`Luxembourg` is the target variable.

```{r}
tar_read(cv_plan)
```

```{r}
tar_read(forecast_plot)
```

```{r}
tar_read(plot_var_imp)
```

```{r}
tar_read(plot_var_resp)
```
```{r}
tar_read(plot_pred_contributions)
```

# Appendix

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{graphicx}

\title{The spread of COVID-19 in a landlocked country the case of
Luxembourg}

\author{
    Bruno Rodrigues
    \thanks{This preprint was written during my free time as a private
citizen, and reflects in no manner the views of the Ministry of Higher
Education and Research, nor the Government of the Grand-Duchy of
Luxembourg.}
   \\
    Statistics department \\
    Ministry of Higher Education and Research, Grand-Duchy of
Luxembourg \\
  Luxembourg, 18-20, Montée de la Pétrusse \\
  \texttt{\href{mailto:bruno.rodrigues@mesr.etat.lu}{\nolinkurl{bruno.rodrigues@mesr.etat.lu}}} \\
  }

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% Pandoc citation processing

\begin{document}
\maketitle

\def\tightlist{}

\begin{abstract}
Enter the text of your abstract here.
\end{abstract}

\keywords{
    blah
   \and
    blee
   \and
    bloo
   \and
    these are optional and can be removed
  }

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

In this article, I posit the following hypothesis: due to its quite
unique characteristics, the spread of COVID-19 in a landlocked country
like Luxembourg is the exact opposite of the spread of COVID-19 that can
be observed on an island country such as New Zealand, or Madagascar. A
landlocked country like Luxembourg, which is furthermore highly
dependent on foreign workers, has many more difficulties to control the
spread of COVID-19 within its borders. Unlike an island country, a
landlocked country that is highly tied to its neighbours cannot simply
close its borders and put a very hard lockdown in place to control the
pandemic. Or if the landlocked country does that, as soon as it opens
its borders, the disease will start spreading again. To illustrate this
idea, I will discuss how COVID-19 starting spreading, but not only
within the borders of Luxembourg, but rather within the so-called
Greater Region. The Greater Region \emph{a space for cross-border
cooperation in the heart of Europe} and is composed of the Grand-Duchy
of Luxembourg, two Belgian Provinces, two French Départements and two
German Bundesländer.

@@

The approach I will be using in this paper is thus as follows: I will
train a machine learning model to predict the spread of COVID-19 in
Luxembourg using openly available data on the weekly positive cases of
COVID-19. However, because of the very tight economic and social
integration of Luxembourg to its neighbours I will use as features
weekly positive cases in the border regions as well as Google Mobility
data\footnote{\url{https://www.google.com/covid19/mobility/}} for
Luxembourg to proxy for hard, and soft, lockdowns. I will show that lags
of weekly cases in the neighbouring regions predict cases for
Luxembourg. The end goal however, is \emph{not} to build a model to
predict how many weekly positive cases will be detected in Luxembourg.
This would be a fools errand; in order to predict the future, the future
has to look like the past, but in the case of this pandemic there is
absolutely no guarantee that the future will look like the past, and
there are many reasons for this. First of all, people are constantly
adapting their behaviour, and public health policies are also constantly
being tuned, and getting sometimes more restrictive, sometimes more
relaxed. Secondly, vaccines have started being administrated and it
would be impossible to predict the effect on weekly positive cases using
the approach I'm using. Finally, there's also the threat of the
variants. Here again, it is impossible to predict which new variants
could arise and how much more contagious -and deadly- these could be. So
then, why bother with this paper? The end goal is not prediction, but
explainability. Once the model is trained, I will use explainability
methods to show which variables, and their interaction with each other,
predict positive cases for Luxembourg. This will be a clear illustration
of the hypothesis that I posited at the beginning; that a landlocked
country like Luxembourg which is very tightly economically and socially
integrated with its neighbours cannot fight a pandemic on its own, but
must cooperate with its neighbours. This argument can also be applied to
any other country in the same situation as Luxembourg or even to the
constituent states of a federal nation. Unfortunately, the virus does
not respect the borders of sovereign nations.

\hypertarget{the-covid-19-pandemic-in-the-greater-region}{%
\section{The COVID-19 pandemic in the Greater
Region}\label{the-covid-19-pandemic-in-the-greater-region}}

COVID-19 started to spread in the Greater Region around end of February.
Two french \emph{départements}, the \emph{Bas-Rhin} and the
\emph{Haut-Rhin}, which together form the historical region of Alsace,
were hit very hard. A religious gathering in the \emph{Haut-Rhin}, with
more than 2000 worshipers is the very likely start of the spread in
Alsace, and then to the neighbouring Lorraine administrative region.
Alsace is part of the \emph{Grand Est} region, as is Lorraine. Two
\emph{départements} of \emph{Lorraine} are part of the Greater Region,
and the disease spread there quickly as well.

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(epid\_curves)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/epid_curves-1.pdf}
\caption{Test}
\end{figure}

In Figure @ref(fig:epid\_curves)

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(epidem\_map)}\SpecialCharTok{$}\NormalTok{map\_first\_wave}
\end{Highlighting}
\end{Shaded}

\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-1-1}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(epidem\_map)}\SpecialCharTok{$}\NormalTok{map\_second\_wave}
\end{Highlighting}
\end{Shaded}

\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-2-1}

\hypertarget{the-data-used}{%
\section{The data used}\label{the-data-used}}

Data on positive cases from the regions of the Greater Region was
collected through each of the countries' open data portal. The levels of
detail were heterogeneous, with Belgium and Germany providing a high
level of detail (daily cases by sex, age group, Province in the case of
Belgium, and Land- and Stadtkreise in the case of Germany), followed by
France (daily cases by department and age group), with Luxembourg
providing the least amount of details; only daily cases at the national
level. In order to simplify the process of getting the data from all
these sources, we wrote an R package called
\texttt{\{covidGrandeRegion\}} which can be found on the following
\href{https://github.com/b-rodrigues/covidGrandeRegion}{github
repository}. This R package provides several functions to download daily
or weekly data, either for one single country or for the whole of the
Greater Region as well as a function to call an interactive map of the
region with a timeline, making it easy to visualise the spread of the
disease through the region. It is also possible to normalize the data by
dividing the daily or weekly cases by the size of the population in each
sub-region. Another variable that was included comes from the
\href{https://www.google.com/covid19/mobility/}{Google Mobility
website}. This data shows on a daily basis how communities move since
the start of the pandemic. This data is used here as proxy for
lockdowns. The plot below shows the daily percentage change in time
spent at home in Luxembourg:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(plot\_mobility)}
\end{Highlighting}
\end{Shaded}

\includegraphics{paper_files/figure-latex/unnamed-chunk-3-1.pdf}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{print\_data }\OtherTok{\textless{}{-}} \FunctionTok{tar\_read}\NormalTok{(data\_for\_model) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\FunctionTok{across}\NormalTok{(}\FunctionTok{where}\NormalTok{(is.numeric), }\SpecialCharTok{\textasciitilde{}}\FunctionTok{round}\NormalTok{(., }\DecValTok{0}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{t}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{as.data.frame}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{  tibble}\SpecialCharTok{::}\FunctionTok{rownames\_to\_column}\NormalTok{(}\AttributeTok{var=}\StringTok{"Variable names"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{select}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{6}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{unite}\NormalTok{(}\StringTok{"First 5 observations"}\NormalTok{, }\DecValTok{2}\SpecialCharTok{:}\DecValTok{6}\NormalTok{, }\AttributeTok{sep =} \StringTok{", "}\NormalTok{) }

\FunctionTok{kbl}\NormalTok{(print\_data, }\AttributeTok{booktabs =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{position =} \StringTok{"center"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{tabular}[t]{ll}
\toprule
Variable names & First 5 observations\\
\midrule
week & 2020-02-24, 2020-03-02, 2020-03-09, 2020-03-16, 2020-03-23\\
Luxembourg & 0,   0,   0, 112, 164\\
lag\_Belgique\_01 & 0,    0,    3,   11,   33\\
lag\_Belgique\_02 & 0,    0,    0,    3,   11\\
lag\_Belgique\_03 & 0,    0,    0,    0,    3\\
\addlinespace
lag\_Belgique\_04 & 0,    0,    0,    0,    0\\
lag\_Deutschland\_01 & 0,   1,   3,  21,  37\\
lag\_Deutschland\_02 & 0,   0,   1,   3,  21\\
lag\_Deutschland\_03 & 0,   0,   0,   1,   3\\
lag\_Deutschland\_04 & 0,   0,   0,   0,   1\\
\addlinespace
lag\_France\_01 & 0,   0,   2,  43,  39\\
lag\_France\_02 & 0,   0,   0,   2,  43\\
lag\_France\_03 & 0,   0,   0,   0,   2\\
lag\_France\_04 & 0,   0,   0,   0,   0\\
lag\_stay\_home\_01 & 0,  1,  1,  4,  4\\
\addlinespace
lag\_stay\_home\_02 & 0,  0,  1,  1,  4\\
lag\_stay\_home\_03 & 0,  0,  0,  1,  1\\
lag\_stay\_home\_04 & 0,  0,  0,  0,  1\\
\bottomrule
\end{tabular}

\texttt{Luxembourg} is the target variable.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(cv\_plan)}
\end{Highlighting}
\end{Shaded}

\includegraphics{paper_files/figure-latex/unnamed-chunk-5-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(forecast\_plot)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in max(ids, na.rm = TRUE): no non-missing arguments to max; returning
## -Inf
\end{verbatim}

\includegraphics{paper_files/figure-latex/unnamed-chunk-6-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(plot\_var\_imp)}
\end{Highlighting}
\end{Shaded}

\includegraphics{paper_files/figure-latex/unnamed-chunk-7-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(plot\_var\_resp)}
\end{Highlighting}
\end{Shaded}

\includegraphics{paper_files/figure-latex/unnamed-chunk-8-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{tar\_read}\NormalTok{(plot\_pred\_contributions)}
\end{Highlighting}
\end{Shaded}

\includegraphics{paper_files/figure-latex/unnamed-chunk-9-1.pdf}

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\bibliographystyle{unsrt}
\bibliography{references.bib}

\end{document}
